---
title: "R Notebook"
output: html_notebook
---

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
```


# --- LOAD & PREPARE ----------------------------------------------------------

```{r}


tidy    <- read.csv("all_experiments_tidy.csv",    stringsAsFactors = FALSE)
summary <- read.csv("all_experiments_summary.csv", stringsAsFactors = FALSE)

exp_order <- c(
  "Exp1_Adaptive_Busiest_Short",
  "Exp2_Stubborn_Busiest_Short",
  "Exp3_Adaptive_Random_Short",
  "Exp4_Stubborn_Random_Short",
  "Exp5_Adaptive_Busiest_Long",
  "Exp6_Stubborn_Busiest_Long",
  "Exp7_Adaptive_Random_Long",
  "Exp8_Stubborn_Random_Long"
)

exp_labels <- c(
  "Exp1_Adaptive_Busiest_Short" = "Exp1\nAdaptive·Busiest·Short",
  "Exp2_Stubborn_Busiest_Short" = "Exp2\nStubborn·Busiest·Short",
  "Exp3_Adaptive_Random_Short"  = "Exp3\nAdaptive·Random·Short",
  "Exp4_Stubborn_Random_Short"  = "Exp4\nStubborn·Random·Short",
  "Exp5_Adaptive_Busiest_Long"  = "Exp5\nAdaptive·Busiest·Long",
  "Exp6_Stubborn_Busiest_Long"  = "Exp6\nStubborn·Busiest·Long",
  "Exp7_Adaptive_Random_Long"   = "Exp7\nAdaptive·Random·Long",
  "Exp8_Stubborn_Random_Long"   = "Exp8\nStubborn·Random·Long"
)

tidy <- tidy %>%
  mutate(
    experiment        = factor(experiment, levels = exp_order),
    never_adapted     = as.logical(never_adapted),
    cartel_type       = factor(cartel_type,       levels = c("Adaptive", "Stubborn")),
    enforcement_target   = factor(enforcement_target,   levels = c("Busiest", "Random")),
    enforcement_duration = factor(enforcement_duration, levels = c("Short", "Long")),
    # Recode first_yellow_tick: -1 means never adapted
    first_yellow_tick_clean = ifelse(first_yellow_tick <= 0, NA, first_yellow_tick)
  )

summary <- summary %>%
  mutate(
    experiment        = factor(experiment, levels = exp_order),
    cartel_type       = factor(cartel_type,       levels = c("Adaptive", "Stubborn")),
    enforcement_target   = factor(enforcement_target,   levels = c("Busiest", "Random")),
    enforcement_duration = factor(enforcement_duration, levels = c("Short", "Long"))
  )

# Shared color scales
col_cartel   <- c("Adaptive" = "#2196F3", "Stubborn" = "#FF5722")
col_strategy <- c("Busiest"  = "#9C27B0", "Random"   = "#78909C")
col_duration <- c("Short"    = "#43A047", "Long"     = "#F9A825")

theme_rq <- theme_minimal(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "#f4f4f4", color = NA),
    panel.spacing    = unit(1, "lines"),
    legend.position  = "bottom",
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(size = 12, color = "grey40")
  )
```


# =============================================================================
# RQ1 — Does tolerance matter under smart pressure?
# Exp 1 vs Exp 2 | Factor: cartel type | Fixed: Busiest · Short
# =============================================================================

```{r}

rq1_data <- tidy %>%
  filter(enforcement_target == "Busiest", enforcement_duration == "Short")

p_rq1 <- rq1_data %>%
  select(experiment, cartel_type,
         drug_through_yellow_rate_pct,
         seizure_rate_pct,
         count_links_with_color_yellow) %>%
  pivot_longer(
    cols      = c(drug_through_yellow_rate_pct, seizure_rate_pct,
                  count_links_with_color_yellow),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = recode(metric,
    "drug_through_yellow_rate_pct"   = "FDI (%)",
    "seizure_rate_pct"               = "Seizure rate (%)",
    "count_links_with_color_yellow"  = "Yellow links"
  )) %>%
  ggplot(aes(x = cartel_type, y = value, fill = cartel_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +
  scale_fill_manual(values = col_cartel) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title    = "RQ1 — Does loss tolerance matter under intelligent enforcement?",
    subtitle = "Exp 1 vs Exp 2 · Busiest Transit · Short duration · diamond = mean",
    x        = "Cartel type",
    y        = NULL,
    fill     = "Cartel type"
  ) +
  theme_rq

#ggsave("RQ1_tolerance_smart_pressure.png", p_rq1, width = 12, height = 5, dpi = 300)
p_rq1
```

# =============================================================================
# RQ2 — What triggers the Balloon Effect most?
# FDI across Exp 1, 2, 3, 5 to isolate strategy, duration, adaptation
# =============================================================================
```{r}

rq2_data <- tidy %>%
  filter(experiment %in% c(
    "Exp1_Adaptive_Busiest_Short",
    "Exp2_Stubborn_Busiest_Short",
    "Exp3_Adaptive_Random_Short",
    "Exp5_Adaptive_Busiest_Long",
    "Exp6_Stubborn_Busiest_Long"
  )) %>%
  mutate(
    exp_label = recode(as.character(experiment),
      "Exp1_Adaptive_Busiest_Short" = "Exp1\nAdaptive·Busiest·Short\n(baseline)",
      "Exp2_Stubborn_Busiest_Short" = "Exp2\nStubborn·Busiest·Short\n(↓ adaptation)",
      "Exp3_Adaptive_Random_Short"  = "Exp3\nAdaptive·Random·Short\n(↓ intelligence)",
      "Exp5_Adaptive_Busiest_Long"  = "Exp5\nAdaptive·Busiest·Long\n(↑ duration)",
      "Exp6_Stubborn_Busiest_Long"  = "Exp6\nStubborn·Busiest·Long\n(↑ duration)",
    ),
    exp_label = factor(exp_label, levels = c(
      "Exp1\nAdaptive·Busiest·Short\n(baseline)",
      "Exp2\nStubborn·Busiest·Short\n(↓ adaptation)",
      "Exp3\nAdaptive·Random·Short\n(↓ intelligence)",
      "Exp5\nAdaptive·Busiest·Long\n(↑ duration)",
      "Exp6\nStubborn·Busiest·Long\n(↑ duration)"
    ))
  )

p_rq2 <- ggplot(rq2_data, aes(x = exp_label, y = drug_through_yellow_rate_pct,
                               fill = enforcement_target)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +
  scale_fill_manual(values = col_strategy) +
  labs(
    title    = "RQ2 — What triggers the Balloon Effect most?",
    subtitle = "One variable changed at a time from Exp1 baseline · diamond = mean",
    x        = NULL,
    y        = "FDI — traffic on shadow routes (%)",
    fill     = "Enforcement strategy"
  ) +
  theme_rq

#ggsave("RQ2_balloon_effect_trigger.png", p_rq2, width = 12, height = 6, dpi = 300)
p_rq2

```

# =============================================================================
# RQ3 — Is it better to be smart or persistent?
# Exp 5 vs Exp 7 | Factor: strategy | Fixed: Adaptive · Long
# =============================================================================
```{r}


rq3_data <- tidy %>%
  filter(cartel_type == "Adaptive", enforcement_duration == "Long")

p_rq3 <- rq3_data %>%
  select(experiment, enforcement_target,
         drug_through_yellow_rate_pct,
         seizure_rate_pct,
         total_delivered_value) %>%
  pivot_longer(
    cols      = c(drug_through_yellow_rate_pct, seizure_rate_pct,
                  total_delivered_value),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = recode(metric,
    "drug_through_yellow_rate_pct" = "FDI (%)",
    "seizure_rate_pct"             = "Seizure rate (%)",
    "total_delivered_value"        = "Total deliveries"
  )) %>%
  ggplot(aes(x = enforcement_target, y = value, fill = enforcement_target)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +
  scale_fill_manual(values = col_strategy) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title    = "RQ3 — Is it better to be smart or persistent?",
    subtitle = "Exp 5 vs Exp 7 · Adaptive cartel · Long duration · diamond = mean",
    x        = "Enforcement strategy",
    y        = NULL,
    fill     = "Enforcement strategy"
  ) +
  theme_rq

#ggsave("RQ3_smart_vs_persistent.png", p_rq3, width = 12, height = 5, dpi = 300)
p_rq3
```

# =============================================================================
# RQ4 — Does rigidity matter when enforcement is blind?
# Exp 3 vs Exp 4 | Factor: cartel type | Fixed: Random · Short
# =============================================================================

```{r}

rq4_data <- tidy %>%
  filter(enforcement_target == "Random", enforcement_duration == "Short")

p_rq4 <- rq4_data %>%
  select(experiment, cartel_type,
         drug_through_yellow_rate_pct,
         seizure_rate_pct,
         count_links_with_color_yellow) %>%
  pivot_longer(
    cols      = c(drug_through_yellow_rate_pct, seizure_rate_pct,
                  count_links_with_color_yellow),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = recode(metric,
    "drug_through_yellow_rate_pct"   = "FDI (%)",
    "seizure_rate_pct"               = "Seizure rate (%)",
    "count_links_with_color_yellow"  = "Yellow links"
  )) %>%
  ggplot(aes(x = cartel_type, y = value, fill = cartel_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +
  scale_fill_manual(values = col_cartel) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title    = "RQ4 — Does rigidity matter when enforcement is blind?",
    subtitle = "Exp 3 vs Exp 4 · Random Transit · Short duration · diamond = mean",
    x        = "Cartel type",
    y        = NULL,
    fill     = "Cartel type"
  ) +
  theme_rq

#ggsave("RQ4_rigidity_random_enforcement.png", p_rq4, width = 12, height = 5, dpi = 300)
p_rq4

```


# =============================================================================
# RQ5 — Can persistence break organizational rigidity?
# Exp 2 vs Exp 6 | Factor: duration | Fixed: Stubborn · Busiest
# Two panels: FDI + first yellow tick distribution
# =============================================================================
```{r}
rq5_data <- tidy %>%
  filter(cartel_type == "Stubborn", enforcement_target == "Busiest")

# Panel A — FDI and seizure rate
p_rq5a <- rq5_data %>%
  select(experiment, enforcement_duration,
         drug_through_yellow_rate_pct, seizure_rate_pct) %>%
  pivot_longer(
    cols      = c(drug_through_yellow_rate_pct, seizure_rate_pct),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = recode(metric,
    "drug_through_yellow_rate_pct" = "FDI (%)",
    "seizure_rate_pct"             = "Seizure rate (%)"
  )) %>%
  ggplot(aes(x = enforcement_duration, y = value, fill = enforcement_duration)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +
  scale_fill_manual(values = col_duration) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title    = "RQ5 — Can persistence break organizational rigidity?",
    subtitle = "Exp 2 vs Exp 6 · Stubborn cartel · Busiest Transit · diamond = mean",
    x        = "Enforcement duration",
    y        = NULL,
    fill     = "Duration"
  ) +
  theme_rq

# Panel B — first yellow tick (when adaptation eventually fires)
p_rq5b <- rq5_data %>%
  filter(!is.na(first_yellow_tick_clean)) %>%
  ggplot(aes(x = first_yellow_tick_clean, fill = enforcement_duration)) +
  geom_histogram(binwidth = 30, alpha = 0.8, color = "white", position = "identity") +
  scale_fill_manual(values = col_duration) +
  facet_wrap(~ enforcement_duration, ncol = 1) +
  labs(
    title    = "RQ5 — Tick of first shadow route activation (stubborn cartel only)",
    subtitle = "Runs where cartel never adapted are excluded",
    x        = "Tick of first yellow link",
    y        = "Count of runs",
    fill     = "Duration"
  ) +
  theme_rq

#ggsave("RQ5a_persistence_rigidity.png",   p_rq5a, width = 10, height = 5, dpi = 300)
#ggsave("RQ5b_first_yellow_tick.png",      p_rq5b, width = 8,  height = 6, dpi = 300)
p_rq5a
p_rq5b
```


# =============================================================================
# RQ6 — Best vs worst enforcement configuration
# Exp 1 vs Exp 8 | All factors differ simultaneously
# =============================================================================

```{r}
rq6_data <- tidy %>%
  filter(experiment %in% c(
    "Exp1_Adaptive_Busiest_Short",
    "Exp8_Stubborn_Random_Long"
  )) %>%
  mutate(config = recode(as.character(experiment),
    "Exp1_Adaptive_Busiest_Short" = "Exp1: Best config\n(Adaptive · Busiest · Short)",
    "Exp8_Stubborn_Random_Long"   = "Exp8: Worst config\n(Stubborn · Random · Long)"
  ))

p_rq6 <- rq6_data %>%
  select(config, drug_through_yellow_rate_pct, seizure_rate_pct,
         total_delivered_value, count_links_with_color_yellow) %>%
  pivot_longer(
    cols      = c(drug_through_yellow_rate_pct, seizure_rate_pct,
                  total_delivered_value, count_links_with_color_yellow),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = recode(metric,
    "drug_through_yellow_rate_pct"   = "FDI (%)",
    "seizure_rate_pct"               = "Seizure rate (%)",
    "total_delivered_value"          = "Total deliveries",
    "count_links_with_color_yellow"  = "Yellow links"
  )) %>%
  ggplot(aes(x = config, y = value,
             fill = config)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +
  scale_fill_manual(values = c(
    "Exp1: Best config\n(Adaptive · Busiest · Short)" = "#9C27B0",
    "Exp8: Worst config\n(Stubborn · Random · Long)"  = "#78909C"
  )) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  labs(
    title    = "RQ6 — Best vs worst enforcement configuration",
    subtitle = "Exp 1 vs Exp 8 · All three factors differ · diamond = mean",
    x        = NULL,
    y        = NULL,
    fill     = "Configuration"
  ) +
  theme_rq +
  theme(axis.text.x = element_blank())

#ggsave("RQ6_best_vs_worst.png", p_rq6, width = 11, height = 8, dpi = 300)
p_rq6

```

# =============================================================================
# RQ7 — Does the cartel ever die?
# All 8 experiments · deliveries + seizure rate + pct never adapted
# =============================================================================
```{r}
# Panel A — total deliveries across all experiments (network never fully collapses)
p_rq7a <- ggplot(tidy, aes(x = experiment, y = total_delivered_value,
                             fill = cartel_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.3, size = 1.2) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +
  scale_fill_manual(values = col_cartel) +
  scale_x_discrete(labels = exp_labels) +
  labs(
    title    = "RQ7 — Does the cartel ever die?",
    subtitle = "Total deliveries across all runs · the network never fully collapses · diamond = mean",
    x        = NULL,
    y        = "Total deliveries at tick 1000",
    fill     = "Cartel type"
  ) +
  theme_rq +
  theme(axis.text.x = element_text(size = 9))

# Panel B — % of runs where cartel never adapted (from summary)
p_rq7b <- summary %>%
  ggplot(aes(x = experiment, y = pct_never_adapted, fill = cartel_type)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_text(aes(label = paste0(round(pct_never_adapted, 1), "%")),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = col_cartel) +
  scale_x_discrete(labels = exp_labels) +
  scale_y_continuous(limits = c(0, 110), breaks = seq(0, 100, 25)) +
  labs(
    title    = "RQ7 — Share of runs where cartel never built shadow routes",
    subtitle = "100% = cartel never adapted in any run · 0% = every run had at least one yellow link",
    x        = NULL,
    y        = "% of runs: never adapted",
    fill     = "Cartel type"
  ) +
  theme_rq +
  theme(axis.text.x = element_text(size = 9))

#ggsave("RQ7a_cartel_deliveries.png",  p_rq7a, width = 13, height = 6, dpi = 300)
#ggsave("RQ7b_never_adapted_pct.png",  p_rq7b, width = 13, height = 6, dpi = 300)
p_rq7a
p_rq7b

```



