---
title: "R Notebook"
output: html_notebook
---

# =============================================================================
# visualizations.R
# Requires: all_experiments_tidy.csv and all_experiments_summary.csv
# =============================================================================

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(kableExtra)
library(knitr)
```


# --- LOAD DATA ---------------------------------------------------------------


```{r}
# FIXED: Using read.csv because reshape script now saves with commas
tidy    <- read.csv("all_experiments_tidy.csv",    stringsAsFactors = FALSE)
summary <- read.csv("all_experiments_summary.csv", stringsAsFactors = FALSE)

exp_order <- c(
  "Exp1_Adaptive_Busiest_Short",
  "Exp2_Stubborn_Busiest_Short",
  "Exp3_Adaptive_Random_Short",
  "Exp4_Stubborn_Random_Short",
  "Exp5_Adaptive_Busiest_Long",
  "Exp6_Stubborn_Busiest_Long",
  "Exp7_Adaptive_Random_Long",
  "Exp8_Stubborn_Random_Long"
)

# FIXED: Unified preparation into one object 'tidy_prepared'
tidy_prepared <- tidy %>%
  mutate(
    experiment = factor(experiment, levels = exp_order),
    plot_title = case_when(
      grepl("Exp1", experiment) ~ "E1-Adaptive",
      grepl("Exp2", experiment) ~ "E2-Stubborn",
      grepl("Exp3", experiment) ~ "E3-Adaptive",
      grepl("Exp4", experiment) ~ "E4-Stubborn",
      grepl("Exp5", experiment) ~ "E5-Adaptive",
      grepl("Exp6", experiment) ~ "E6-Stubborn",
      grepl("Exp7", experiment) ~ "E7-Adaptive",
      grepl("Exp8", experiment) ~ "E8-Stubborn"
    ),
    enforcement_duration = factor(enforcement_duration, 
                                 levels = c("Short", "Long"),
                                 labels = c("Enforcement duration: Short", 
                                            "Enforcement duration: Long"))
  )

summary$experiment <- factor(summary$experiment, levels = exp_order)
```


# =============================================================================
# PLOT 1 — Seizure Rate by Experiment
# Dataset: tidy
# =============================================================================

```{r}

p1 <- ggplot(tidy_prepared, aes(x = enforcement_target, y = seizure_rate_pct, fill = enforcement_target)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c("Busiest" = "#9C27B0", "Random" = "#607D8B")) +
  facet_grid(plot_title ~ enforcement_duration, scales = "free_y") + 
  stat_summary(fun = mean, geom = "label", 
               aes(label = sprintf("%.2f%%", ..y..)),
               fill = "white", alpha = 0.8, size = 3, fontface = "bold") +
  labs(title = "Seizure Rate Analysis", y = "Seizure Rate (%)", fill = "Strategy") +
  theme_minimal()

p1

# After your ggplotly(p) or just p code:

ggsave("Seizure Rate analysis.png", 
     plot = p1,               # 'p' is your ggplot object
      width = 12,             # Width in inches
       height = 10,            # Height in inches
       dpi = 300,              # High resolution for "beauty"
      limitsize = FALSE)
```


# =============================================================================
# PLOT 2 — FDI by Experiment
# Dataset: tidy
# =============================================================================

```{r}
# FIXED: Column name changed to drug_through_yellow_rate_pct
p_fdi <- ggplot(tidy_prepared, aes(x = enforcement_target, y = drug_through_yellow_rate_pct, fill = enforcement_target)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.3, size = 1) +
  stat_summary(fun = mean, geom = "label", 
               aes(label = sprintf("%.2f%%", ..y..)),,
               fill = "white", alpha = 0.8, size = 3, fontface = "bold") +
  scale_fill_manual(values = c("Busiest" = "#9C27B0", "Random" = "#607D8B")) +
  facet_grid(plot_title ~ enforcement_duration, scales = "free_y") + 
  labs(title = "FDI Analysis (Drug Flow Displacement)", y = "FDI (% through shadow links)") +
  theme_minimal()

p_fdi

#3. Save it to lock dimensions and prevent "crumpling"
ggsave("FDI_Analysis_Grid.png", 
       plot = p_fdi, 
       width = 12, 
       height = 10, 
       dpi = 300)

p_fdi
```


# =============================================================================
# PLOT 3 — Deliveries vs Seizures side by side
# Dataset: tidy
# =============================================================================

```{r}
# FIXED: Reference tidy_prepared and corrected column names
# --- 4. PLOT 3: Deliveries vs Seizures (With Value Labels) -------------------

p_ds <- tidy_prepared %>%
  select(experiment, total_delivered_value, total_seized_value) %>%
  # Pivot to long format so we can plot both metrics side-by-side
  pivot_longer(
    cols      = starts_with("total_"), 
    names_to  = "metric", 
    values_to = "value"
  ) %>%
  mutate(metric = recode(metric, 
    "total_delivered_value" = "Deliveries", 
    "total_seized_value"    = "Seizures"
  )) %>%
  ggplot(aes(x = experiment, y = value, fill = metric)) +
  # 1. The Bars
  stat_summary(fun = "mean", geom = "bar", position = position_dodge(0.8), alpha = 0.8) +
  
  # 2. NEW: The Numerical Labels
  # We use vjust = -0.5 to nudge the text slightly above the top of the bar
  stat_summary(
    fun = "mean", 
    geom = "text", 
    aes(label = sprintf("%.1f", ..y..)), 
    position = position_dodge(width = 0.8), 
    vjust = -0.5, 
    size = 3.5, 
    fontface = "bold"
  ) +
  
  scale_fill_manual(values = c("Deliveries" = "#4CAF50", "Seizures" = "purple")) +
  labs(
    title = "Average Deliveries vs Seizures by Experiment", 
    subtitle = "Labels indicate the mean volume for each metric",
    x = NULL, 
    y = "Average Volume",
    fill = "Metric"
  ) +
  # Expand the y-axis slightly so the labels don't get cut off at the top
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

# Display the plot
p_ds

# Save the plot
ggsave("Deliveries_vs_Seizures_Labeled.png", plot = p_ds, width = 12, height = 8, dpi = 300)
```


# =============================================================================
# PLOT 4 — Heatmap: Mean FDI across the factorial grid
# Dataset: summary
# =============================================================================

This plot represents a Heatmap Analysis of Flow Displacement, showing how successfully different cartel types bypass enforcement under various conditions.

```{r}
p_meanfdi <- summary |>
  ggplot(aes(x = enforcement_target, y = cartel_type, fill = mean_drug_through_yellow_rate_pct)) +
  geom_tile(color = "white", linewidth = 1.5) +
  # Change the label logic here to add the % sign
  geom_text(aes(label = paste0(round(mean_drug_through_yellow_rate_pct, 1), "%")),
            size = 5, 
            fontface = "bold") +
  facet_wrap(~ enforcement_duration) +
  scale_fill_gradient(low = "#E3F2FD", high = "#DC6826") +
  labs(
    title    = "Mean FDI — Factorial Grid",
    subtitle = "Faceted by enforcement duration (Short vs Long)",
    x        = "Enforcement Strategy",
    y        = "Cartel Type",
    fill     = "Mean FDI"
  ) +
  theme_minimal(base_size = 13)

p_meanfdi

ggsave("Mean FDI.png", 
     plot = p_meanfdi, 
      width = 12, 
      height = 10, 
       dpi = 300)


```


# =============================================================================
# PLOT 5 — Interaction plot: FDI by cartel type x enforcement strategy
# Dataset: summary
# =============================================================================

```{r}
##SKIPPARE
# FIXED: mean_fdi -> mean_drug_through_yellow_rate_pct
summary %>%
  filter(enforcement_duration == "Short") %>%
  ggplot(aes(x = enforcement_target, y = mean_drug_through_yellow_rate_pct, 
             color = cartel_type, group = cartel_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = mean_drug_through_yellow_rate_pct - sd_drug_through_yellow_rate_pct, 
                    ymax = mean_drug_through_yellow_rate_pct + sd_drug_through_yellow_rate_pct), width = 0.1) +
  scale_color_manual(values = c("Adaptive" = "#2196F3", "Stubborn" = "#FF5722")) +
  labs(title = "Interaction Plot: FDI Strategy Response", y = "Mean FDI (%)") +
  theme_minimal()
```


# =============================================================================
# PLOT 6 — Shadow route accumulation (yellow links at end of run)
# Dataset: tidy
# =============================================================================

```{r}
# --- PLOT 6: Shadow Route Accumulation (Updated Layout) ----------------------

p_shadow <- ggplot(tidy_prepared, aes(x = enforcement_target, y = count_links_with_color_yellow, fill = enforcement_target)) +
  # Boxplot and Jitter
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.3, size = 1) +
  
  # Mean labels (formatted to 1 decimal place since this is a count)
  stat_summary(fun = mean, geom = "label", 
               aes(label = sprintf("%.1f", ..y..)),
               fill = "white", alpha = 0.8, size = 3, fontface = "bold") +
  
  # Use the same color palette as Plot 1 & 2
  scale_fill_manual(values = c("Busiest" = "#9C27B0", "Random" = "#607D8B")) +
  
  # Facet by Experiment (Rows) and Duration (Columns)
  facet_grid(plot_title ~ enforcement_duration, scales = "free_y") + 
  
  labs(
    title = "Shadow Route Accumulation Analysis", 
    subtitle = "Count of yellow (clandestine) links at end of 1000 ticks",
    x = "Targeting Strategy",
    y = "Yellow Links (Count)",
    fill = "Strategy"
  ) +
  theme_minimal()

# Save with standard "beauty" dimensions
ggsave("Shadow_Route_Analysis_Grid.png", 
      plot = p_shadow, 
      width = 12, 
     height = 10, 
       dpi = 300)

p_shadow

```
  

# =============================================================================
# PLOT 7 — FDI vs Clustering Coefficient (Duijn et al. test)
# Dataset: tidy
# =============================================================================

```{r}

## SKIPPARE ##

# FIXED: Column names synced with tidy dataset
ggplot(tidy_prepared, aes(x = drug_through_yellow_rate_pct, y = clustering_rate_pct, 
                          color = plot_title, shape = cartel_type)) +
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group = 1), method = "lm", color = "black", linetype = "dashed") +
  labs(title = "FDI vs Clustering Coefficient", x = "FDI (%)", y = "Clustering CoefFficient (%)") +
  theme_minimal()
```



```{r}

# 1. Prepare and format the data
table_data_long <- summary %>%
  # Ensure Adaptive comes before Stubborn for clean grouping
  arrange(cartel_type, experiment) %>% 
  rename(
    "Type" = cartel_type,
    "Experiment" = experiment,
    "Target" = enforcement_target,
    "Duration" = enforcement_duration,
    "Runs" = n_runs,
    "Avg Delivered" = mean_delivered,
    "Avg Seized" = mean_seized,
    "Seizure Rate" = mean_seizure_rate_pct,
    "FDI (Flow %)" = mean_drug_through_yellow_rate_pct,
    "Clustering" = mean_clustering_rate_pct
  ) %>%
  select(Type, Experiment, Target, Duration, Runs, 
         `Avg Delivered`, `Avg Seized`, `Seizure Rate`, `FDI (Flow %)`, `Clustering`)

# 2. Generate one single styled table
kable(table_data_long, 
      caption = "Summary Statistics: All Experiments",
      booktabs = TRUE, 
      digits = 2,
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, 
                position = "left") %>%
  # This merges the "Adaptive" and "Stubborn" cells so they don't repeat every row
  collapse_rows(columns = 1, valign = "top") %>% 
  row_spec(0, , color = "white", background = "#4A4A4A") %>%
  column_spec(2,  ) # Bold the Experiment name
```


# =============================================================================
# PLOT 8 — Highlighting inefficiency of random strategy
# Dataset: summary
# =============================================================================


```{r}

library(ggplot2)
library(dplyr)

# 1. Filter the summary data for the specific experiments
# We use case_when to create clean labels for the X-axis
plot_data <- summary %>%
  filter(grepl("Exp1_|Exp3_|Exp7_", experiment)) %>%
  mutate(
    display_label = case_when(
      grepl("Exp3", experiment) ~ "Exp 3\n(Random / Short)",
      grepl("Exp7", experiment) ~ "Exp 7\n(Random / Long)",
      grepl("Exp1", experiment) ~ "Exp 1\n(Targeted / Short)"
    )
  )

# 2. Set the factor order so they appear side-by-side as requested
plot_data$display_label <- factor(plot_data$display_label, 
                                 levels = c("Exp 3\n(Random / Short)", 
                                            "Exp 7\n(Random / Long)", 
                                            "Exp 1\n(Targeted / Short)"))

# 3. Create the high-contrast bar plot
p_comparison <- ggplot(plot_data, aes(x = display_label, y = mean_drug_through_yellow_rate_pct, fill = display_label)) +
  # Using mean_seizure_rate_pct based on your requirement
  geom_col(aes(y = mean_seizure_rate_pct), color = "black", width = 0.6) +
  # Add the percentage labels on top of the bars
  geom_text(aes(y = mean_seizure_rate_pct, label = sprintf("%.1f%%", mean_seizure_rate_pct)), 
            vjust = -0.8, size = 6, fontface = "bold") +
  # High-contrast color palette: Greys for Random, Red for Targeted
  scale_fill_manual(values = c(
    "Exp 3\n(Random / Short)" = "#FFBF7F", # Light Grey
    "Exp 7\n(Random / Long)"  = "#A5EDFF", # Dark Grey
    "Exp 1\n(Targeted / Short)" = "#B2FF8C"  
  )) +
  # Formatting axes
  scale_y_continuous(limits = c(0, 25), expand = c(0, 0)) +
  labs(
    title = "Targeting Efficiency vs. Brute Force",
    subtitle = "Comparing Seizure Rates: Why Hub-Targeting beats Long-Duration Random policing",
    x = "",
    y = "Mean Seizure Rate (%)",
    caption = "Exp 1 (5 days targeted) > Exp 7 (25 days random)"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", size = 20),
    axis.text.x = element_text(face = "bold", color = "black", size = 12),
    axis.title.y = element_text(margin = margin(r = 10))
    
  )

# Display and Save
print(p_comparison)
ggsave("Strategy_Efficiency_Comparison.png", p_comparison, width = 10, height = 7, dpi = 300)

```

# =============================================================================
# PLOT 9 - Efficiency paradox
# Dataset: summary
# =============================================================================



```{r}

library(ggplot2)
library(dplyr)

plot_data <- data.frame(
  Label = factor(
    c("Exp3\nAdaptive·Random·Short", "Exp2\nStubborn·Busiest·Short",
      "Exp1\nAdaptive·Busiest·Short", "Exp6\nStubborn·Busiest·Long",
      "Exp5\nAdaptive·Busiest·Long"),
    levels = c("Exp3\nAdaptive·Random·Short", "Exp2\nStubborn·Busiest·Short",
               "Exp1\nAdaptive·Busiest·Short", "Exp6\nStubborn·Busiest·Long",
               "Exp5\nAdaptive·Busiest·Long")
  ),
  Clustering = c(3.29, 5.53, 11.82, 10.50, 10.56),
  Seizure    = c(2.35, 22.74, 19.94, 40.92, 41.64),
  Phase = factor(
    c("Baseline", "Short Raid", "Short Raid", "Long Surge", "Long Surge"),
    levels = c("Baseline", "Short Raid", "Long Surge")
  )
)

# Scale seizure onto the clustering y-axis
scale_factor <- 52 / 16   # right axis max / left axis max

p <- ggplot(plot_data, aes(x = Label)) +
  # Bars — clustering coefficient
  geom_col(aes(y = Clustering, fill = Phase),
           color = "black", linewidth = 0.4, width = 0.55) +
  # Clustering labels above bars
  geom_text(aes(y = Clustering, label = paste0(round(Clustering, 1), "%")),
            vjust = -0.5, size = 3.3, color = "grey40", fontface = "bold") +
  # Line — seizure rate (scaled to left axis)
  geom_line(aes(y = Seizure / scale_factor, group = 1),
            color = "#1565C0", linewidth = 1.5) +
  geom_point(aes(y = Seizure / scale_factor),
             color = "#1565C0", size = 4, shape = 21,
             fill = "#1565C0", stroke = 1.5) +
  # Seizure labels near dots
  geom_text(aes(y = Seizure / scale_factor,
                label = paste0(round(Seizure, 1), "%")),
            vjust = -1.1, size = 3.3, color = "#1565C0", fontface = "bold") +
  # Dual y-axis
  scale_y_continuous(
    name     = "Clustering coefficient (%)",
    limits   = c(0, 16),
    expand   = expansion(mult = c(0, 0.12)),
    sec.axis = sec_axis(~ . * scale_factor,
                        name   = "Seizure rate (%)",
                        breaks = seq(0, 50, 10),
                        labels = function(x) paste0(x, "%"))
  ) +
  scale_fill_manual(values = c(
    "Baseline"   = "#E3F2FD",
    "Short Raid" = "#F5CA63",
    "Long Surge" = "#DC6826"
  )) +
  labs(
    title    = "The Efficiency-Hardening Paradox",
    subtitle = "Bars: clustering coefficient  |  Line + dots: seizure rate",
    x        = NULL,
    fill     = "Phase"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y.right = element_text(color = "#1565C0"),
    axis.text.y.right  = element_text(color = "#1565C0"),
    axis.text.x        = element_text(size = 9),
    plot.title         = element_text(face = "bold"),
    legend.position    = "bottom"
  )

print(p)
ggsave("Enforcement_Tradeoff_DualAxis.png", p, width = 11, height = 6, dpi = 300)

```

